---
title: "relationship_pursuit_study"
author: Milena Chobanyan
format: html
editor: visual
html:
  theme: flatly
  toc: TRUE
  toc-title: "Report Sections"
  toc-depth: 2
  toc-location: left
  number-selections: FALSE
  code-fold: TRUE
  code-summary: "Show code"
  code-tools: TRUE
  df-print: paged
  highlight-style: tango
  embed-resources: TRUE
  smooth-scroll: TRUE
---

## Preparing Report

```{r}
#| label: setup
#| echo: false
#| message: false
#| warning: false

# Set global chunk options for the entire document
knitr::opts_chunk$set(
  echo = TRUE,      # display code in output
  message = FALSE,  # suppress package messages
  warning = FALSE   # suppress warnings
)

# Prevent automatic conversion of strings to factors
options(stringsAsFactors = FALSE)  # keep character variables as text
```

## Introduction

The goal is to convert the raw data-set into an analysis-ready dataset by loading data, defining variables, cleaning responses, and constructing a scale score.

## Software Installations

```{r}
#| label: packages
#| echo: true
#| message: false
#| warning: false

library(tidyverse)
library(psych)  
library(afex) 
library(emmeans) 
library(svglite)
```

## Uploading Data

```{r}
data <- read.csv("data/relationship_pursuit_copy.csv")
```

```{r}
data <- data %>%
  rename(check = Q8)
```

```{r}
names(data)
```

```{r}
# filter out ppl who didn't answer commit questions
data <- data %>%
  filter(if_all(check:condition, ~ !is.na(.)))
```

```{r}
head(data)
dim(data)
```

```{r}
names(data)
```

## Variable Preparation and Scale Construction

##### Define the Study Variable List

```{r}
#| label: define_vars
#| echo: true
#| message: false
#| warning: false

# Create a vector of variables included in the study
study_vars <- c(
  "condition",
  "check",
  "sex",       # participant sex
  "age",       # participant age
  "Commit_1","Commit_2","Commit_3","Commit_4","Commit_5","Commit_6","Commit_7","Commit_8",  # Commitment items
  "Long_Term_Traits_1","Long_Term_Traits_2","Long_Term_Traits_3","Long_Term_Traits_4","Long_Term_Traits_5","Long_Term_Traits_6","Long_Term_Traits_7","Long_Term_Traits_8","Long_Term_Traits_9","Long_Term_Traits_10",  # Long-term items
  "Short_Term_Traits_1","Short_Term_Traits_2","Short_Term_Traits_3","Short_Term_Traits_4","Short_Term_Traits_5","Short_Term_Traits_6","Short_Term_Traits_7","Short_Term_Traits_8","Short_Term_Traits_9","Short_Term_Traits_10" # Short-term items
)

# Check the number of variables in the list
length(study_vars)  # confirms total number matches expected count
```

##### Creating analysis dataset

```{r}
#| label: prepare_data
#| echo: true
#| message: false
#| warning: false

# Convert all variables to numeric format
data <- data %>% #and then
  mutate(across(Long_Term_Traits_1:Commit_8, as.numeric))  # ensures numeric type for all variables

# Summarize dataset to confirm conversion
summary(data)  # checks values and missing data - remove for final report
```

**Label the experimental condition**

```{r}
#| label: label_condition
#| echo: true
#| message: false
#| warning: false

# Convert condition variable to labeled factor
table(data$condition)
```

```{r}
#| label: label_sex
#| echo: true
#| message: false
#| warning: false

# Convert condition variable to labeled factor
data <- data %>%
  mutate(
    sex = factor(
      sex,                     # original numeric variable
      levels = c(1, 2, 3),             # original coding
      labels = c("Other", "Female", "Other")  # readable labels
    )
  )

# Verify condition counts
data %>%
  count(sex)  # counts observations in each condition
```

------------------------------------------------------------------------

##### Creating Participant ID

```{r}
#| label: create_id
#| echo: true
#| message: false
#| warning: false

# Create a unique ID for each participant
data <- data %>%
  mutate(id = row_number())  # assigns sequential ID values

# Preview the ID variable
data %>% select(id) %>% head()  # confirms ID structure
```

##### Reverse Coding

```{r}
#| label: reverse_code_Commit
#| echo: true
#| message: false
#| warning: false

# Reverse-code selected commitment items
data <- data %>%
  mutate(
    Commit_2r  = 8 - Commit_2,   # reverse-codes Commit_2
    Commit_4r  = 8 - Commit_4,   # reverse-codes Commit_4
    Commit_6r = 8 - Commit_6,  # reverse-codes Commit_6
    Commit_8r = 8 - Commit_8   # reverse-codes Commit_8
  )

# Compare original and reversed values
data %>%
  select(Commit_4, Commit_4r) %>%  # selects variables for comparison
  head()  # displays first rows
```

```{r}
#| label: reverse_code_Long_Term_Traits
#| echo: true
#| message: false
#| warning: false

# Reverse-code selected long term items so that higher score means higher emphasis on trustworthiness
data <- data %>%
  mutate(
    Long_Term_Traits_3r  = 8 - Long_Term_Traits_3,   # reverse-codes Long_Term_Traits3
    Long_Term_Traits_4r  = 8 - Long_Term_Traits_4,   # reverse-codes Long_Term_Traits4
    Long_Term_Traits_6r = 8 - Long_Term_Traits_6,  # reverse-codes Long_Term_Traits6
    Long_Term_Traits_7r = 8 - Long_Term_Traits_7,   # reverse-codes Long_Term_Traits7
    Long_Term_Traits_10r = 8 - Long_Term_Traits_10 # reverse-codes Long_Term_Traits10
  )

# Compare original and reversed values
data %>%
  select(Long_Term_Traits_3, Long_Term_Traits_3r) %>%  # selects variables for comparison
  head()  # displays first rows
```

```{r}
#| label: reverse_code_Short_Term_Traits
#| echo: true
#| message: false
#| warning: false

# Reverse-code selected short term items so that higher score means higher emphasis on trustworthiness
data <- data %>%
  mutate(
    Short_Term_Traits_3r  = 8 - Short_Term_Traits_3,   # reverse-codes Short_Term_Traits3
    Short_Term_Traits_4r  = 8 - Short_Term_Traits_4,   # reverse-codes Short_Term_Traits4
    Short_Term_Traits_6r = 8 - Short_Term_Traits_6,  # reverse-codes Short_Term_Traits6
    Short_Term_Traits_7r = 8 - Short_Term_Traits_7,   # reverse-codes Short_Term_Traits7
    Short_Term_Traits_10r = 8 - Short_Term_Traits_10 # reverse-codes Short_Term_Traits10
  )

# Compare original and reversed values
data %>%
  select(Short_Term_Traits_3, Short_Term_Traits_3r) %>%  # selects variables for comparison
  head()  # displays first rows
```

##### Defining Scales

```{r}
#| label: define_scale_Commit
#| echo: true
#| message: false
#| warning: false

# Define commitment scale items
Commit_items <- c(
  "Commit_1","Commit_2r","Commit_3","Commit_4r","Commit_5","Commit_6r","Commit_7","Commit_8r"
)

# Check that all items exist in dataset
setNames(Commit_items %in% names(data), Commit_items)  # verifies item presence
```

```{r}
#| label: mutate_Trait_NAs
#| echo: true
#| message: false
#| warning: false

data <- data %>%
  mutate(across(contains("Traits"), ~coalesce(., 0)))
```

```{r}
#| label: mutate_combine_Traits1-10
#| echo: true
#| message: false
#| warning: false

data <- data %>%
  mutate(
    Trait_1_Total  = Long_Term_Traits_1  + Short_Term_Traits_1,
    Trait_2_Total  = Long_Term_Traits_2  + Short_Term_Traits_2,
    Trait_3_Total  = Long_Term_Traits_3  + Short_Term_Traits_3,
    Trait_4_Total  = Long_Term_Traits_4  + Short_Term_Traits_4,
    Trait_5_Total  = Long_Term_Traits_5  + Short_Term_Traits_5,
    Trait_6_Total  = Long_Term_Traits_6  + Short_Term_Traits_6,
    Trait_7_Total  = Long_Term_Traits_7  + Short_Term_Traits_7,
    Trait_8_Total  = Long_Term_Traits_8  + Short_Term_Traits_8,
    Trait_9_Total  = Long_Term_Traits_9  + Short_Term_Traits_9,
    Trait_10_Total = Long_Term_Traits_10 + Short_Term_Traits_10
  )
```

```{r}
#| label: define_warmth_items
#| echo: true
#| message: false
#| warning: false

# Define commitment scale items
Warmth_items <- c(
  "Trait_1_Total", "Trait_2_Total", "Trait_5_Total", "Trait_8_Total", "Trait_9_Total"
)

# Check that all items exist in dataset
setNames(Warmth_items %in% names(data), Warmth_items)  # verifies item presence
```

```{r}
#| label: define_attractiveness_items
#| echo: true
#| message: false
#| warning: false

# Define commitment scale items
Attractiveness_items <- c(
  "Trait_3_Total", "Trait_4_Total", "Trait_6_Total", "Trait_7_Total", "Trait_10_Total"
)

# Check that all items exist in dataset
setNames(Attractiveness_items %in% names(data), Attractiveness_items)  # verifies item presence
```

##### Computing Scale Scores and Scale Variables

```{r}
#| label: compute_scale_Commit
#| echo: true
#| message: false
#| warning: false

# Compute commitment scale as the row mean of items
data <- data %>%
  mutate(
    Commit_scale = rowMeans(as.matrix(across(all_of(Commit_items))), na.rm = TRUE)  # calculates scale score
  )

# Check proportion of missing scale scores
mean(is.na(data$Commit_scale))  # should be close to 0

# Check distribution and range of the scale
summary(data$Commit_scale)  # should fall within expected range
```

```{r}
#| label: compute_scale_Wamrth
#| echo: true
#| message: false
#| warning: false

# Compute long term scale as the row mean of items
data <- data %>%
  mutate(
    Warmth_scale = rowMeans(as.matrix(across(all_of(Warmth_items))), na.rm = TRUE)  # calculates scale score
  )

# Check proportion of missing scale scores
mean(is.na(data$Warmth_scale))  # should be close to 0

# Check distribution and range of the scale
summary(data$Warmth_scale)  # should fall within expected range
```

```{r}
#| label: compute_scale_Attractiveness
#| echo: true
#| message: false
#| warning: false

# Compute long term scale as the row mean of items
data <- data %>%
  mutate(
    Attractiveness_scale = rowMeans(as.matrix(across(all_of(Attractiveness_items))), na.rm = TRUE)  # calculates scale score
  )

# Check proportion of missing scale scores
mean(is.na(data$Attractiveness_scale))  # should be close to 0

# Check distribution and range of the scale
summary(data$Attractiveness_scale)  # should fall within expected range
```

##### Final Check

```{r}
#| label: final_check
#| echo: true
#| message: false
#| warning: false

# Check dataset dimensions
dim(data)  # confirms number of rows and columns

# Check final variable names
names(data)  # confirms variables available for next steps
```

##### Computing Median of Each Scale

```{r}
#| label: compute_Commit_median
#| echo: true
#| warning: false
#| message: false

# Compute the median value of the Commit scale
median(data$Commit_scale, na.rm = TRUE)
```

```{r}
#| label: compute_Warmth_median
#| echo: true
#| warning: false
#| message: false

# Compute the median value of the Short Term scale
median(data$Warmth_scale, na.rm = TRUE)
```

```{r}
#| label: compute_Attractiveness_median
#| echo: true
#| warning: false
#| message: false

# Compute the median value of the Short Term scale
median(data$Attractiveness_scale, na.rm = TRUE)
```

##### Creating Split Variables

```{r}
#| label: create_Commit_split
#| echo: true
#| warning: false
#| message: false

# Compute the median of the Commit scale
Commit_median <- median(data$Commit_scale, na.rm = TRUE)

# Add a new categorical variable based on the median split
data <- data %>%
  mutate(
    # Classify scores at or below the median as "Low" and above the median as "High"
    Commit_msplit = if_else(Commit_scale <= Commit_median, "Low", "High") %>%
      # Convert to a factor with "Low" as the reference level
      factor(levels = c("Low", "High"))
  )
```

```{r}
#| label: create_Warmth_split
#| echo: true
#| warning: false
#| message: false

# Compute the median of the Short Term scale
Warmth_median <- median(data$Warmth_scale, na.rm = TRUE)

# Add a new categorical variable based on the median split
data <- data %>%
  mutate(
    # Classify scores at or below the median as "Low" and above the median as "High"
    Warmth_msplit = if_else(Warmth_scale <= Warmth_median, "Low", "High") %>%
      # Convert to a factor with "Low" as the reference level
      factor(levels = c("Low", "High"))
  )
```

```{r}
#| label: create_Attractiveness_split
#| echo: true
#| warning: false
#| message: false

# Compute the median of the Short Term scale
Attractiveness_median <- median(data$Attractiveness_scale, na.rm = TRUE)

# Add a new categorical variable based on the median split
data <- data %>%
  mutate(
    # Classify scores at or below the median as "Low" and above the median as "High"
    Attractiveness_msplit = if_else(Attractiveness_scale <= Attractiveness_median, "Low", "High") %>%
      # Convert to a factor with "Low" as the reference level
      factor(levels = c("Low", "High"))
  )
```

##### Inspecting Split Classifications

```{r}
#| label: check_Commit_split
#| echo: true
#| warning: false
#| message: false

# Preview the continuous Commit scores alongside the new group labels
data %>%
  select(Commit_scale, Commit_msplit) %>%  # Keep only the scale and split variables
  head(10) %>%                      # Show the first 10 rows
  knitr::kable()                   # Format as a table for display

# Count the number of participants in each Commit group
data %>%
  count(Commit_msplit) %>%            # Count observations in each group
  knitr::kable()                  # Format as a table for display
```

```{r}
#| label: check_Warmth_split
#| echo: true
#| warning: false
#| message: false

# Preview the continuous Short Term scores alongside the new group labels
data %>%
  select(Warmth_scale, Warmth_msplit) %>%  # Keep only the scale and split variables
  head(10) %>%                      # Show the first 10 rows
  knitr::kable()                   # Format as a table for display

# Count the number of participants in each Short Term group
data %>%
  count(Warmth_msplit) %>%            # Count observations in each group
  knitr::kable()                  # Format as a table for display
```

```{r}
#| label: check_Attractiveness_split
#| echo: true
#| warning: false
#| message: false

# Preview the continuous Short Term scores alongside the new group labels
data %>%
  select(Attractiveness_scale, Attractiveness_msplit) %>%  # Keep only the scale and split variables
  head(10) %>%                      # Show the first 10 rows
  knitr::kable()                   # Format as a table for display

# Count the number of participants in each Short Term group
data %>%
  count(Attractiveness_msplit) %>%            # Count observations in each group
  knitr::kable()                  # Format as a table for display
```

##### **Compute internal consistency (alpha and omega)**

```{r}
# Create item-only data frames for each scale
Commit_item_df <- data %>% select(all_of(Commit_items))
Warmth_item_df <- data %>% select(all_of(Warmth_items))
Attractiveness_item_df <- data %>% select(all_of(Attractiveness_items))

# Compute internal consistency estimates (store results)
Commit_alpha <- psych::alpha(Commit_item_df)
Commit_omega <- psych::omega(Commit_item_df, plot = FALSE)

Warmth_alpha <- psych::alpha(Warmth_item_df)
Warmth_omega <- psych::omega(Warmth_item_df, plot = FALSE)

Attractiveness_alpha <- psych::alpha(Attractiveness_item_df)
Attractiveness_omega <- psych::omega(Attractiveness_item_df, plot = FALSE)

# Print only alpha and omega coefficients
tibble(
  scale = c("Commit", "Warmth", "Attractiveness"),
  alpha = c(
    Commit_alpha$total$raw_alpha,
    Warmth_alpha$total$raw_alpha,
    Attractiveness_alpha$total$raw_alpha
  ),
  omega = c(
    Commit_omega$omega.tot,
    Warmth_alpha$total$raw_alpha,
    Attractiveness_alpha$total$raw_alpha
  )
)
```

***In the output, locate and record:***

- ***Cronbach’s alpha (α)***
- ***McDonald’s omega (ω)***

------------------------------------------------------------------------

**Descriptive Statistics (Continuous Variables)**

```{r}
#| label: descriptives_continuous
#| echo: true
#| warning: false
#| message: false

# Descriptive statistics for continuous scale variables
psych::describe(
  data %>% select(Commit_scale, Warmth_scale, Attractiveness_scale)
) %>%
  knitr::kable(digits = 2)
```

**Run descriptives for categorical variables**

*Categorical variables are summarized using counts rather than means.*

```{r}
#| label: descriptives_categorical_condition
#| echo: true
#| warning: false
#| message: false

# Count participants by condition
data %>%
  count(condition) %>%
  knitr::kable()
```

**Convert counts to percentages**

*Percentages make group composition easier to interpret than raw counts.*

```{r}
#| label: categorical_percentages_condition
#| echo: true
#| warning: false
#| message: false

# Convert counts to percentages
data %>%
  count(condition) %>%
  mutate(percent = round(100 * n / sum(n), 1)) %>%
  knitr::kable()
```

**Run the ANOVA for DV**

*Mood Condition is coded as 0 = Negative and 1 = Positive, and Neuroticism is coded as Low vs. High, as defined during data preparation.*

```{r}
#| label: anova_model1
#| echo: true
#| warning: false
#| message: false

# 2x2 between-subjects ANOVA predicting Warmth scores
# DV: Warmth_scale
# Factors: condition and Commit_msplit
a1_Warmth <- afex::aov_ez(
  data = data,
  id = "id",
  dv = "Warmth_scale",
  between = c("condition", "Commit_msplit")
)

# Print the ANOVA table
a1_Warmth
```

```{r}
#| label: anova_model2
#| echo: true
#| warning: false
#| message: false

# 2x2 between-subjects ANOVA predicting Attractiveness scores
# DV: Attractiveness_scale
# Factors: condition and Commit_msplit
a2_Attractiveness <- afex::aov_ez(
  data = data,
  id = "id",
  dv = "Attractiveness_scale",
  between = c("condition", "Commit_msplit")
)

# Print the ANOVA table
a2_Attractiveness
```

*Identify the F statistic, numerator and denominator degrees of freedom, p-value, and generalized eta-squared for each effect.*

------------------------------------------------------------------------

### Inspect marginal (overall) means and confidence intervals (Main effects / “group means”)

*In this step, you will examine group means and confidence intervals to determine the direction of effects tested in the ANOVA.*

```{r}
#| label: marginal_means
#| echo: true
#| warning: false
#| message: false

# Marginal (group) means for mood condition
# Averaged across levels of neuroticism
emmeans(a1_Warmth, ~ condition)

# Marginal (group) means for neuroticism
# Averaged across levels of mood condition
emmeans(a1_Warmth, ~ Commit_msplit)
```

```{r}
#| label: marginal_means2
#| echo: true
#| warning: false
#| message: false

# Marginal (group) means for mood condition
# Averaged across levels of neuroticism
emmeans(a2_Attractiveness, ~ condition)

# Marginal (group) means for neuroticism
# Averaged across levels of mood condition
emmeans(a2_Attractiveness, ~ Commit_msplit)
```

***Locate the marginal (overall) means and confidence intervals.***

------------------------------------------------------------------------

### Inspect cell means and confidence intervals (Interaction Means)

```{r}
#| label: cell_means
#| echo: true
#| warning: false
#| message: false

# Raw estimated marginal means
emmeans(a1_Warmth, ~ condition * Commit_msplit)
```

```{r}
#| label: cell_means2
#| echo: true
#| warning: false
#| message: false

# Raw estimated marginal means
emmeans(a2_Attractiveness, ~ condition * Commit_msplit)
```

Inspect the output: all condition × group combinations are present

means and confidence intervals are displayed

------------------------------------------------------------------------

## **Bar Plot**

This plot displays model-estimated mean depression scores across mood condition, shown separately for low and high neuroticism.

```{r}
#| label: interaction_plot_bar1
#| echo: true
#| warning: false
#| message: false

# Traitsx interaction plot (bar + 95% CI)
# This figure shows model-estimated mean depression scores across mood condition, separately for low vs. high Commitment, with 95% confidence intervals.

p_Warmth_bar <- ggplot(
  as.data.frame(emmeans(a1_Warmth, ~ condition * Commit_msplit)),  # cell means from the ANOVA model
  aes(
    x = condition,        # x-axis = experimental condition
    y = emmean,          # y-axis = estimated mean depression score
    fill = Commit_msplit    # fill color = commitment levels
  )
) +
  geom_col(
    position = position_dodge(width = 0.65),  # place Low vs. High side-by-side
    width = 0.55,                              # width of each bar
    color = "white"                            # white outlines for visual separation
  ) +
  geom_errorbar(
    aes(
      ymin = lower.CL,   # lower bound of 95% CI
      ymax = upper.CL    # upper bound of 95% CI
    ),
    position = position_dodge(width = 0.65),  # align error bars with bars
    width = 0.12,                              # width of error bar caps
    linewidth = 0.7                            # thickness of error bars
  ) +
  labs(
    x = "Condition",        # x-axis label
    y = "Mean Trait Score", # y-axis label
    fill = "Commitment"         # legend title
  ) +
  scale_fill_brewer(
    palette = "Dark2"            # colorblind-friendly fill colors
  ) +
  scale_y_continuous(
    breaks = 1:7                 # show all response scale values
  ) +
  coord_cartesian(
    ylim = c(1, 7)               # display full 1–7 scale without clipping bars
  ) +
  theme_classic(
    base_size = 14               # clean, publication-style theme
  )

print(p_Warmth_bar)
```

```{r}
#| label: interaction_plot_bar2
#| echo: true
#| warning: false
#| message: false

# Traitsx interaction plot (bar + 95% CI)
# This figure shows model-estimated mean depression scores across mood condition, separately for low vs. high Commitment, with 95% confidence intervals.

p_Attractiveness_bar <- ggplot(
  as.data.frame(emmeans(a2_Attractiveness, ~ condition * Commit_msplit)),  # cell means from the ANOVA model
  aes(
    x = condition,        # x-axis = experimental condition
    y = emmean,          # y-axis = estimated mean depression score
    fill = Commit_msplit    # fill color = commitment levels
  )
) +
  geom_col(
    position = position_dodge(width = 0.65),  # place Low vs. High side-by-side
    width = 0.55,                              # width of each bar
    color = "white"                            # white outlines for visual separation
  ) +
  geom_errorbar(
    aes(
      ymin = lower.CL,   # lower bound of 95% CI
      ymax = upper.CL    # upper bound of 95% CI
    ),
    position = position_dodge(width = 0.65),  # align error bars with bars
    width = 0.12,                              # width of error bar caps
    linewidth = 0.7                            # thickness of error bars
  ) +
  labs(
    x = "Condition",        # x-axis label
    y = "Mean Trait Score", # y-axis label
    fill = "Commitment"         # legend title
  ) +
  scale_fill_brewer(
    palette = "Dark2"            # colorblind-friendly fill colors
  ) +
  scale_y_continuous(
    breaks = 1:7                 # show all response scale values
  ) +
  coord_cartesian(
    ylim = c(1, 7)               # display full 1–7 scale without clipping bars
  ) +
  theme_classic(
    base_size = 14               # clean, publication-style theme
  )

print(p_Attractiveness_bar)
```

##### Save Plot

```{r}
#| label: save_figure_warmth
#| echo: true
#| warning: false
#| message: false

# Create figures folder if needed
if (!dir.exists("figures")) dir.create("figures")

# Save whichever plots you want (SVG format for posters)
ggsave("figures/Warmth_bar_plot.svg",   plot = p_Warmth_bar,   width = 7, height = 5)
```

```{r}
#| label: save_figure_attractiveness
#| echo: true
#| warning: false
#| message: false

# Create figures folder if needed
if (!dir.exists("figures")) dir.create("figures")

# Save whichever plots you want (SVG format for posters)
ggsave("figures/Attractiveness_bar_plot.svg",   plot = p_Attractiveness_bar,   width = 7, height = 5)
```
